---
title: "Dplyr tut"
author: "Ziqi Liu"
date: "May 10, 2018"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

# 1.0 Loading dplyr

- 5 basic verbs:
1. filter()
2. select()
3. arrange()
4. mutate()
5. summarise()
``` {r setup}
suppressMessages(library(dplyr))
library(hflights)
sessionInfo()

#explore data
data(hflights)
head(hflights)
```

- tbl_df creates a local data frame, wraps text so prints nicely
``` {r local_data_frame}
flights <- tbl_df(hflights)
flights
```

# 2.0 filter()

- allows you to filter by rows matching certain criteria

### Command Structure
- first argument is a data frame, following arguments are what to filter by
- return value is a data frame
``` {r filter}
filter(flights, Month==1, DayofMonth==1)
```

# 3.0 select()

- allows you to pick columns by name

### Command Structure
- first argument is a data frame, following arguments are column names
``` {r select}
select(flights, DepTime, ArrTime, FlightNum)
```

## 3.1 Chaining or Pipelining

- allows you to perform multiple operations in one line by nesting commands
- use the %>% operator (pronounced "and then")
``` {r chain}
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  filter(DepDelay > 60)

# use dataframe flights AND THEN select column names UniqueCarrier and DepDelay AND THEN filter by rows where DepDelay value > 60
```

# 4.0 arrange()

- allows you to reorder rows by sorting in ascending order (default)

### Command Structure

- call dataframe, then select columns, then arrange by the column of choice
- for descending order use arrange(desc(column_name))
``` {r arrange}
flights %>%
  select(UniqueCarrier, DepDelay) %>%
  arrange(DepDelay)

# selects UniqueCarrier and DepDelay columns and then rearranges by DepDelay ascending
```

# 5.0 mutate()

- allows you to add new variables that are functions of existing variables
``` {r mutate}
flights %>%
  select(Distance, AirTime) %>%
  mutate(Speed = Distance/AirTime*60)

# creates new variable Speed that is a function of Distance and Airtime

# to add to existing dataframe, don't select the columns
flights <- flights %>%
  mutate(Speed = Distance/AirTime*60)
```

# 6.0 summarise()

- reduces variables to values, useful for data that has been grouped by one or more variables
- group_by() creates the groups that will be operated on, summarise() uses the provided aggregation function to summarise each group
``` {r summarise}
flights %>%
  group_by(Dest) %>%
  summarise(avg_delay = mean(ArrDelay, na.rm=TRUE))

# create a table grouped by Dest, then summarise each group into avg_delay by taking the mean of ArrDelay
```

# Notes
- glimpse() lets you see the dataframe data better

``` {r huh}
library(rgl)
yes <- flights %>% select(AirTime, Speed, Distance)
print(plot3d(yes))
```

